---
title: "Data471 EDA Project"
author: "Group 21: Loren Whithair, Nancy Grace Busto, Hamid Mehmood, Leon Cai"
date: "`r Sys.Date()`"
output:
  pdf_document:
    md_extensions: -implicit_figures
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = "center")
```

```{=latex}
\setlength{\LTcapwidth}{\linewidth}
```

#### Additional Authors

The students I collaborated with in Phase 1 are:

- Nancy Grace Busto (300 691 941)
- Hamid Mehmood (300 693 052)
- Leon Cai (300 692 424)



# Executive Summary

TODO

# Background

Reasonable access to food sources has a significant impact a person's 
quality of life. One key component of accessibility is the distance a person is
required to travel in order to access affordable and healthy food, such as ingredients
from a supermarket. This is a commonly discussed phenomenon in the United States of America,
and is referred to as "food deserts".

This impact of longer distances is exacerbated for those with less financial resources,
such as individuals in poverty. These individuals have less resources to overcome the geographical
barrier to access, such as the money required for extra fuel, or the ability to accumulate savings to buy in bulk.

The Food Access Research Atlas (FARA) provides various measures of food access in the United States
by Census Tract, in combination with data regarding income, poverty rates, age and ethnicity breakdowns. 
Census Tracts are geographical regions defined such that they roughly equally divide the population.

The aim of this analysis is to understand in greater detail the characteristics of Census Tracts
that are considered to have a significant number of individuals that both suffer from low income and low access to food.

Specifically, I aim to answer the question:

> _How do the **demographic characteristics** of a Census Tract relate to whether it is considered to be an area that suffers from **low income** and **low access**?_


Addressing this question can achieve the following:

1. It may highlight unforeseen consequences, or potentially hidden biases, in decisions made by those that 
influence the location of supermarkets, such as local governments and supermarket businesses
2. It may identify areas and population groups that future initiatives should prioritise, to have the greatest impact
<!-- 1-3 pages -->



# Description of the FARA dataset


We used the 2019 Large Retailer Access Map (LRAM) data, part of the United
States Department of Agriculture's Food Access Research Atlas (FARA,
USDA ERS 2019 -- see References). This data is available directly from the USDA ERS
[LRAM data download](https://www.ers.usda.gov/media/5627/2019-large-retailer-access-map-lram-formerly-known-as-the-food-access-research-atlas-fara-data.zip?v=29938).

This data describes, by Census Tract, the accessibility of supermarkets to the population of that Census Tract. Food Access is defined in multiple variables by several distances - for example, **LAPOP1_10** describes the number of people that are more than 1 mile from a supermarket for an Urban Census Tract, and more than 10 miles for a Rural Census Tract. Demographic information such as the population of different ethnicities, the population of children and seniors, is available, both as totals for the census and as the total of each demographic that is more than a specific distance from a food vendor, as measured in the 2010 Census. All 50 US states are covered. 


## Data types and structure

The data contains one row per census tract, with 72,531 rows. Each row also specifies the State and County (character/categorical). The rest of the data exhibits three variable types: binary flags (e.g. **Urban** flags if the Census Tract is an Urban area, **LATracts1** flags if the tract is considered to be low access at a 1 mile distance); tract totals (numeric counts, e.g. **POP2010** for the total population of the tract, **TractKids** for the number of children in the tract); and continuous rates/shares (continuous, e.g. **lapophalfshare** details the share of the tract that are beyond 1/2 mile from a supermarket).

Each population count that represents individuals with insufficient food access, low income, or both, has a corresponding flag. For example, **LAPOP1_10** counts the population of individuals who are more than 1 mile from a supermarket in urban Census Tracts and 10 miles in rural Census Tracts. In comparison, **LA1and10** is a flag to indicate where either at least 500 people, or at least 33%, of the population of that Census Tract are more than 1 mile from a supermarket in urban Census Tracts and 10 miles in rural Census Tracts.


## Variables utilised in this analysis

The following variables were used in this analysis:

TODO: ask Louise how to do this bit exactly...


| Variable Name      | Data Type | Description|
|:-------------------------------------------------------------------|----------:|
| LILATracts_halfAnd10     | Flag (converted to a factor)     | Whether a Census Tract is considered Low Income and Low Access at 1/2 mile for urban Census Tracts, and 10 miles for rural Census Tracts |
| Urban | Flag (converted to a factor) | A flag for whether the Census Tract is Urban (city or densely populated suburb) or Rural | 
| TractBlack_perc | Numerical (between 0 and 100) | The percentage of the population in the Census Tract that is Black |
| TractWhite_perc | Numerical (between 0 and 100) | The percentage of the population in the Census Tract that is White |
| TractHispanic_perc | Numerical (between 0 and 100) | The percentage of the population in the Census Tract that is Black |
| TractAsian_perc | Numerical (between 0 and 100) | The percentage of the population in the Census Tract that is Black |
| TractNHOPI_perc | Numerical (between 0 and 100) | The percentage of the population in the Census Tract that is Black |
| TractAIAN_perc | Numerical (between 0 and 100) | The percentage of the population in the Census Tract that is Black |
| TractOMultir_perc | Numerical (between 0 and 100) | The percentage of the population in the Census Tract that is Black |
| TractBlack_perc | Numerical (between 0 and 100) | The percentage of the population in the Census Tract that is Black |
| TractBlack_perc | Numerical (between 0 and 100) | The percentage of the population in the Census Tract that is Black |
| Population counts and percentages beyond 10 miles from supermarket    | 89.16     |
| Population counts and percentages beyond 20 miles from supermarket     | 97.92     |


 TractAsian_perc * Urban_f
  + TractHispanic_perc #* Urban_f
  + TractNHOPI_perc #* Urban_f
  + TractAIAN_perc * Urban_f
  + TractOMultir_perc #* Urban_f
  + TractBlack_perc 
        # * TractKids_perc 
        # * Urban_f
  + TractAsian_perc 
        # * TractKids_perc 
        * Urban_f
  + TractHispanic_perc 
        * TractKids_perc 
        * Urban_f
        
    
## Completeness

The following columns were fully populated:

- State, County, Urban flag (basic identifier attributes)
- **Pop2010** (total population count of the census tract)
- All flag variables, such as **LA1and10** (flag for low access at 1 mile and 10 miles, urban and rural respectively)

Population tract totals (for example, **TractBlack**, the total number of Black people in the population) have only 0.1% of records missing.

Total population counts for low access exhibited many more missing values - between 20% and 50%, depending on the distance measure. Corresponding flags are zero when the population count is not available.
Variables for low access and low income shared the same percentage of missing values as their low access only counterparts.

Filled records for these fields included measurements of 0, indicating a missing value is conceptually different to the measurement equalling 0.

No clear pattern was found among the missing values. We suspect that these values were left blank if the Census Tract was smaller than the distance described in the variable.

TODO: how were incomplete values handled in this analysis

The response variable for this analysis is a flag, which contains no missing values. Including Census Tracts that are not flagged as low income and low access could have made correlations between demographic information and the outcome weaker. 


# Ethics, Privacy and Security

TODO: revise this in response to feedback

## Ethics

Insights concerning specific demographics, particularly based on ethnicity, have the potential to reinforce biases about those demographics. For example, identifying a correlation between the proportion of Black residents in an area and lower income may contribute to existing stereotypes. Findings in this area must be presented with appropriate contextualisation, and an emphasis on systemic factors, rather than implying outcomes are a direct consequence of the demographic characteristics. 

## Privacy concerns

Every variable we use is a tract-level count, rate, or flag (e.g.
PovertyRate, lablack1share, HUNVFlag) – there are no individual-level records (names, addresses,
or household-level identifiers) anywhere in the dataset, so re-identification of a specific person is not
a realistic risk.

A handful of tracts have a very low population count, and theoretically could be traced back to individuals. Whenever these tracts are used, they will be aggregated into larger groups such as by State or Urban/Rural status, removing any concerns of traceability.

## Security

LRAM data is publicly available for personal use; steps do not need to be taken to secure the data itself.

Analysis and findings can be stored securely in a private Github repository, only granting access to individuals working on the project. Devices that contain local copies of the code can be password-protected. The Github accounts of those with access can be made secure using SSH or other secure server connection protocol.

# Exploratory Data Analysis

```{r data-setup, include=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(patchwork)
library(corrplot)
library(maps)

df <- read.csv(
  "data/2019/2019_Food_Access_Research_Atlas_Data/Food Access Research Atlas.csv",
  stringsAsFactors = FALSE,
  na.strings = c("NA", "NULL", "")
)

num_vars <- c(
  "Pop2010", "OHU2010", "LA1and10", "HUNVFlag", "Urban",
  "PovertyRate", "MedianFamilyIncome", "TractLOWI", "lalowi1share",
  "lawhite1share", "lablack1share", "laasian1share",
  "lanhopi1share", "laaian1share", "laomultir1share", "lahisp1share",
  "lakids1share", "laseniors1share", "TractKids", "TractSeniors",
  "lahunv1share", "lasnap1share", "TractHUNV", "TractSNAP",
  "GroupQuartersFlag"
)
for (v in num_vars) df[[v]] <- as.numeric(df[[v]])

core_vars <- c("LA1and10", "PovertyRate", "MedianFamilyIncome", "Urban", "HUNVFlag")
df <- df[complete.cases(df[, core_vars]), ]

df <- df %>%
  mutate(
    black_share_plain = TractBlack / Pop2010,
    kids_share_plain = TractKids / Pop2010,
    seniors_share_plain = TractSeniors / Pop2010,
    hunv_share_plain = TractHUNV / OHU2010
  )
df$LA1and10_f <- factor(df$LA1and10, labels = c("Not low access", "Low access"))
df$Urban_f <- factor(df$Urban, labels = c("Rural", "Urban"))
df$LILATracts_1And10_f <- factor(df$LILATracts_1And10,
                                  labels = c("Not low income/access", "Low income/access"))
```


```{r cohen-chart, include=FALSE}
#config
path <- "data/2019/2019_Food_Access_Research_Atlas_Data/Food Access Research Atlas.csv" 
boxplotdf <- read.csv(path, stringsAsFactors = FALSE, check.names = FALSE,
               na.strings = c("NA", "NULL", ""))
filter_geography <- "all"      # "all" | "urban" | "rural"
filter_states    <- NULL       # NULL, or c("California", ...)

scope_label <- paste0(
  tools::toTitleCase(filter_geography), " tracts",
  if (is.null(filter_states)) " - all states"
  else paste0(" - ", paste(filter_states, collapse = ", ")))
safe_name <- gsub("[^A-Za-z0-9]+", "_", scope_label)

find_col <- function(data, target) {
  hit <- names(data)[tolower(trimws(names(data))) == tolower(trimws(target))]
  if (length(hit)) hit[1] else NA_character_
}
pop_col   <- find_col(boxplotdf, "POP2010")
ohu_col   <- find_col(boxplotdf, "OHU2010")
urban_col <- find_col(boxplotdf, "Urban")
state_col <- find_col(boxplotdf, "State")

#flags
flag_wanted <- c(
  "LA1and10", "LAhalfand10", "LA1and20",          # (A) low access
  "LowIncomeTracts",                               # (B) low income
  "LILATracts_1And10", "LILATracts_halfAnd10",     # (C) combined
  "LILATracts_1And20")
flag_map <- setNames(lapply(flag_wanted, find_col, data = boxplotdf), flag_wanted)
flag_map <- flag_map[!is.na(flag_map)]
flags    <- unlist(flag_map)

pred_asis     <- c("PovertyRate", "MedianFamilyIncome")
pred_popcount <- c("TractKids", "TractSeniors", "TractWhite", "TractBlack",
                   "TractAsian", "TractHispanic")
pred_hucount  <- c("TractSNAP", "TractHUNV")

keep_meta <- c("CensusTract", state_col, urban_col, pop_col, ohu_col)
keep_meta <- keep_meta[!is.na(keep_meta)]

boxplotdf <- boxplotdf %>%
  select(any_of(c(keep_meta, flags, pred_asis, pred_popcount, pred_hucount)))

#filters
n_before <- nrow(boxplotdf)
if (filter_geography != "all") {
  if (is.na(urban_col)) stop("No 'Urban' column for geography filter.")
  urb  <- suppressWarnings(as.numeric(boxplotdf[[urban_col]]))
  keep <- if (filter_geography == "urban") urb == 1 else urb == 0
  boxplotdf   <- boxplotdf[keep %in% TRUE, ]
}
if (!is.null(filter_states)) {
  if (is.na(state_col)) stop("No 'State' column for state filter.")
  st  <- tolower(trimws(boxplotdf[[state_col]]))
  boxplotdf  <- boxplotdf[st %in% tolower(trimws(filter_states)), ]
}
cat("Scope:", scope_label, "| rows:", n_before, "->", nrow(boxplotdf), "\n")
if (nrow(boxplotdf) == 0) stop("No rows left after filtering.")

#clean
boxplotdf <- boxplotdf %>%
  mutate(across(all_of(c(pred_asis, pred_popcount, pred_hucount,
                         pop_col, ohu_col)), ~ suppressWarnings(as.numeric(.x))),
         across(all_of(flags), ~ suppressWarnings(as.integer(.x))))

pop_vec <- boxplotdf[[pop_col]]; ohu_vec <- boxplotdf[[ohu_col]]
boxplotdf <- boxplotdf %>%
  mutate(across(all_of(pred_popcount), ~ .x / pop_vec, .names = "{.col}_share"),
         across(all_of(pred_hucount),  ~ .x / ohu_vec, .names = "{.col}_share"),
         across(ends_with("_share"),
                ~ ifelse(is.infinite(.x) | is.nan(.x), NA_real_, .x)))

share_cols <- c(paste0(pred_popcount, "_share"), paste0(pred_hucount, "_share"))
predictors <- c(pred_asis, share_cols)

impute_median <- function(x) ifelse(is.na(x), median(x, na.rm = TRUE), x)
boxplotdf <- boxplotdf %>% mutate(across(all_of(predictors), impute_median))

pretty <- c(PovertyRate = "Poverty rate",
            MedianFamilyIncome = "Median family income",
            TractKids_share = "% Kids (0-17)",
            TractSeniors_share = "% Seniors (65+)",
            TractWhite_share = "% White",
            TractBlack_share = "% Black",
            TractAsian_share = "% Asian",
            TractHispanic_share = "% Hispanic",
            TractSNAP_share = "% SNAP households",
            TractHUNV_share = "% No-vehicle households")


pred_groups <- list(
  list(key = "ethnicity", label = "Ethnicity",
       vars = c("TractWhite_share", "TractBlack_share",
                "TractAsian_share", "TractHispanic_share"),
       yfmt = "percent", ncol = 2),
  list(key = "age", label = "Age group",
       vars = c("TractKids_share", "TractSeniors_share"),
       yfmt = "percent", ncol = 2),
  list(key = "households", label = "Households",
       vars = c("TractSNAP_share", "TractHUNV_share"),
       yfmt = "percent", ncol = 2),
  list(key = "income_poverty", label = "Median income & poverty",
       vars = c("MedianFamilyIncome", "PovertyRate"),
       yfmt = "comma", ncol = 2)
)

#box plot function
# range_df (optional): data.frame(predictor, value) with the min/max each
# predictor's y-axis should include, on top of whatever "flag"'s own No/Yes
# data needs. Passing the same range_df to multiple calls (e.g. the three
# ethnicity comparisons below) forces those facets onto matching y-axes,
# so box-height differences between the three figures reflect real
# effect-size differences rather than each figure auto-scaling to itself.
#
# Implemented as one small ggplot per predictor (coord_cartesian sets its
# exact y-range) combined with patchwork, rather than a single faceted
# plot with scales="free_y" -- combining free_y facets with an injected
# geom_blank range and outlier.shape=NA turned out to make every facet's
# range leak into every other facet (a ggplot2 layout quirk), so building
# each panel independently sidesteps that entirely.
# coord_cartesian() zooms the view without recomputing the scale's breaks
# for the new (narrower) window, so ggplot's default breaks (picked for the
# full, un-zoomed data range) can miss the panel's actual visible span
# entirely. Recompute round-number ("nice") breaks from the true visible
# range instead, so each panel still gets clean, evenly-stepped tens/round
# numbers, just scaled to what's actually shown.
nice_breaks <- function(rng) scales::extended_breaks()(rng)

box_by_flag_group <- function(data, flag, family_label, grp, range_df = NULL) {
  y_scale <- if (grp$yfmt == "percent") {
    scale_y_continuous(labels = percent_format(accuracy = 1))
  } else {
    scale_y_continuous(labels = label_comma())
  }

  plot_df <- data %>%
    transmute(.flag = factor(.data[[flag]], levels = c(0, 1),
                             labels = c("No", "Yes")),
              across(all_of(grp$vars))) %>%
    filter(!is.na(.flag)) %>%
    pivot_longer(all_of(grp$vars), names_to = "predictor", values_to = "value") %>%
    mutate(predictor = factor(recode(predictor, !!!pretty),
                              levels = unname(pretty[grp$vars])))

  one_panel <- function(pred_lab) {
    d <- plot_df %>% filter(predictor == pred_lab)
    p <- ggplot(d, aes(.flag, value, fill = .flag)) +
      geom_boxplot(outlier.shape = NA, width = 0.6, linewidth = 0.4) +
      scale_fill_manual(values = c(No = "#965AC5", Yes = "#F4AF98")) +
      # The "%" prefix on labels like "% White" is redundant here -- the
      # y-axis itself is already formatted as a percentage -- so it's
      # dropped from the panel title only (pred_lab itself is left alone,
      # since it's still used as the lookup key into range_df etc. below).
      labs(title = sub("^% ", "", pred_lab), x = NULL, y = NULL) +
      theme_minimal(base_size = 11) +
      theme(
        legend.position = "none",
        plot.title    = element_text(face = "bold", size = 11, hjust = 0.5,
                                     margin = margin(b = 4)),
        axis.text     = element_text(size = 9),
        plot.margin   = margin(8, 10, 8, 10)
      )
    if (is.null(range_df)) {
      p <- p + y_scale
    } else {
      # Fixed to the overall (flag-agnostic) whisker range only -- not
      # unioned with this call's own data -- so the axis is genuinely
      # identical across the three flag comparisons. A flag whose own
      # spread happens to exceed the shared range gets its whiskers
      # clipped at the edges rather than the axis re-expanding, since
      # comparability across figures is the point here.
      target <- range(range_df$value[range_df$predictor == pred_lab], na.rm = TRUE)
      brks <- nice_breaks(target)
      p <- p + coord_cartesian(ylim = target) +
        (if (grp$yfmt == "percent")
           scale_y_continuous(labels = percent_format(accuracy = 1), breaks = brks)
         else
           scale_y_continuous(labels = label_comma(), breaks = brks))
    }
    p
  }

  panels <- lapply(unname(pretty[grp$vars]), one_panel)
  wrap_plots(panels, ncol = grp$ncol) +
  plot_annotation(
    title = paste0(family_label, ":  ", flag),
    theme = theme(
      plot.title = element_text(face = "bold", size = 14, hjust = 0.5)
    )
  )
}

family_of <- function(f) {
  if (grepl("^LILATracts", f)) "LOW INCOME + LOW ACCESS"
  else if (grepl("^LowIncome", f)) "LOW INCOME only"
  else "LOW ACCESS only"
}

# cohen_d function
cohen_d <- function(value, group01) {
  g1 <- value[group01 == 1]; g0 <- value[group01 == 0]
  n1 <- sum(!is.na(g1)); n0 <- sum(!is.na(g0))
  if (n1 < 2 || n0 < 2) return(NA_real_)
  sp <- sqrt(((n1 - 1) * var(g1, na.rm = TRUE) +
              (n0 - 1) * var(g0, na.rm = TRUE)) / (n1 + n0 - 2))
  (mean(g1, na.rm = TRUE) - mean(g0, na.rm = TRUE)) / sp
}

# Long table of d for every flag x predictor, with pretty labels + order.
pred_order   <- unlist(lapply(pred_groups, function(g) g$vars), use.names = FALSE)
pred_order_l <- unname(pretty[pred_order])          # display order (grouped)

d_long <- do.call(rbind, lapply(flags, function(f) {
  data.frame(flag      = unname(f),
             family    = family_of(f),
             predictor = predictors,
             pred_lab  = recode(predictors, !!!pretty),
             d         = vapply(predictors,
                                function(p) cohen_d(boxplotdf[[p]], boxplotdf[[f]]), numeric(1)),
             row.names = NULL, stringsAsFactors = FALSE)
}))
d_long$flag     <- factor(d_long$flag,     levels = unname(flags))
d_long$pred_lab <- factor(d_long$pred_lab, levels = rev(pred_order_l))

# --- Page: heatmap table (numbers printed in each cell) ---------------
lim <- max(abs(d_long$d), na.rm = TRUE)
lim <- if (!is.finite(lim) || lim == 0) 1 else lim
lim <- ceiling(lim / 0.5) * 0.5  # round up to a clean 0.5 step so the legend's outer ticks land on a round number

heat <- ggplot(d_long, aes(flag, pred_lab, fill = d)) +
  geom_tile(colour = NA) +
  geom_text(aes(label = sprintf("%.2f", d)),
            colour = "black", size = 3.4, fontface = "bold") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                       midpoint = 0, limits = c(-lim, lim),
                       breaks = scales::breaks_width(0.5)(c(-lim, lim)),
                       oob = scales::squish, name = NULL) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x   = element_text(angle = 45, hjust = 1, vjust = 1,
                                 size = 9),
    axis.text.y   = element_text(size = 10, hjust = 1),
    axis.ticks    = element_blank(),
    panel.grid    = element_blank(),
    legend.position = "right",
    legend.text   = element_text(hjust = 1),
    plot.margin   = margin(16, 20, 10, 16)
  )

# final plots
get_group <- function(k)
  pred_groups[[which(vapply(pred_groups, function(g) g$key == k, logical(1)))]]
eth_grp <- get_group("ethnicity")
ip_grp  <- get_group("income_poverty")

# Shared y-axis range per ethnicity predictor (see box_by_flag_group's
# range_df argument) -- computed once from the whole dataset, regardless
# of flag, so the Low Access / Low Income / LILA ethnicity figures below
# are directly comparable across pages.
eth_range_df <- do.call(rbind, lapply(eth_grp$vars, function(v) {
  bp <- boxplot.stats(boxplotdf[[v]])$stats
  data.frame(predictor = pretty[[v]], value = c(bp[1], bp[5]))
}))
eth_range_df$predictor <- factor(eth_range_df$predictor,
                                  levels = unname(pretty[eth_grp$vars]))
```

## Exploring Low Access by Ethnicity, Income, and Poverty Rates 

```{r low-access-boxpots-ethnicity, fig.width=5.2, fig.height=2.9}
ff <- unname("LAhalfand10")
box_by_flag_group(boxplotdf, ff, family_of(ff), eth_grp, range_df = eth_range_df)
```
```{=latex}
\begin{center}
```
**Figure 1.1**: Percentages of Census Tract populations that are various ethnicities. Census Tracts are split by Low access flags, measured at a 1/2 mile distance for Urban and 10 miles for Rural. Ethnicities are shown on different scales for visibility
```{=latex}
\end{center}
```

```{r low-access-boxpots-income, fig.width=5.2, fig.height=1.7}
ff <- unname("LAhalfand10")
box_by_flag_group(boxplotdf, ff, family_of(ff), ip_grp)
```

```{=latex}
\begin{center}
```
**Figure 1.2**: Median Family Income and Poverty Rates of Census Tracts, split by Low access flags, measured at a 1/2 mile distance for Urban and 10 miles for Rural. 
```{=latex}
\end{center}
```

Ethnicity shows minimal differences in medians when separating Census Tracts by Low Access status at 1/2 and 10 miles.

The difference in income and poverty rate between Census Tracts by Low Access status is similarly weak: tracts flagged as low access actually have slightly *higher* median income ($70.4k vs $66.1k) and slightly *lower* poverty (11.4% vs 13.0%) than non-flagged tracts. 

## Exploring differences in Urban and Rural Census Tracts

```{r poverty-income-by-urban, fig.width=5.2, fig.height=1.9}

p_income_urban <- ggplot(df, aes(LA1and10_f, MedianFamilyIncome, fill = LA1and10_f)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, linewidth = 0.4) +
  facet_wrap(~ Urban_f) +
  coord_cartesian(ylim = c(0, 165000)) +
  scale_y_continuous(labels = label_comma()) +
  scale_fill_manual(values = c(`Not low access` = "#965AC5", `Low access` = "#F4AF98")) +
  labs(x = NULL, y = "Median family\nincome ($)", title = "MEDIAN FAMILY INCOME:  LA1and10") +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "none",
    plot.title   = element_text(face = "bold", size = 14, hjust = 0.5),
    strip.text   = element_text(face = "bold", size = 11,
                                margin = margin(t = 3, b = 4)),
    axis.text    = element_text(size = 9),
    axis.title.y = element_text(size = 9),
    plot.margin  = margin(8, 10, 8, 10)
  )

p_income_urban
```
```{=latex}
\begin{center}
```
**Figure 2:** Median family income for Urban and Rural census Tracts, where Census Tracts are further split by Low access **Low income** flags, measured at a 1/2 mile distance for Urban and 10 miles for Rural.

```{=latex}
\end{center}
```

In rural tracts, we see that low access tracts exhibit lower median family incomes. 

However, in Urban tracts this effect is reversed. This reversed effect may be caused by densely populated affluent suburban neighbourhoods that do not have a local grocery store. This may explain the lack of noticeable difference in Median Income when grouping Urban and Rural Census Tracts together in Figure 1.2.

## Exploring Low Income and Low Access (LILA) flags by Ethnicity

```{r low-income-low-access-boxpots-ethnicity, fig.width=5.2, fig.height=2.9}
ff <- unname("LILATracts_halfAnd10")
box_by_flag_group(boxplotdf, ff, family_of(ff), eth_grp, range_df = eth_range_df)
```
```{=latex}
\begin{center}
```
**Figure 3**: Percentages of Census Tract populations that are various ethnicities. Census Tracts are split by combined **Low access** and **Low income** flags, measured at a 1/2 mile distance for Urban and 10 miles for Rural. Ethnicities are shown on different scales for visibility
```{=latex}
\end{center}
```

Figure 3 demonstrates there are different relationships between various and the Census Tract's status for Low income and low access (measured at 1/2 mile for Urban and 1 mile for Rural).


The largest ethnicity-based difference between the two statuses is seen in the percentage of the population that is White:  the median percentage for census tracts that are flagged is around 60%, whereas the median percentage for census tracts that are not flagged is closer to 90%.

In contrast, the median percentage of Black individuals in the population for census tracts that are flagged is around 10%, whereas the median percentage of Black individuals in the population for census tracts that are not flagged is less than 5%. This effect is in the opposite direction to the effect of the percentage of White individuals in the population.

We see smaller differences in the percentage of the populations that are Asian and Hispanic between the two statuses, with a lower percentage of Asian individuals in Census Tracts considered low income and low access compared to census tracts without this status, 
and a higher percentage of Hispanic individuals in Census Tracts considered low income and low access.

Differences emerge in demographic percentages in Figure 3 that are not present in Figure 1.1, implying that the observed effect may be largely explained by the low income discriminating feature.


# Detailed Data Analysis

```{r detailed_analysis_setup}

#----------------------
set.seed(1899)

all.data <- read.csv(
  "data/2019/2019_Food_Access_Research_Atlas_Data/Food Access Research Atlas.csv",
  stringsAsFactors = FALSE,
  na.strings = c("NA", "NULL", "")
)

all.data <- as_tibble(all.data)

# Convert CensusTract to character:
all.data$CensusTract <- as.character(all.data$CensusTract)

# Convert Urban/Rural to factor, with Urban as reference
all.data$Urban_f <- factor(all.data$Urban, levels=c(1,0), labels = c("Urban", "Rural"))


# Response variable
all.data$LILATracts_halfAnd10_f <- factor(all.data$LILATracts_halfAnd10,
                                    levels=c(0,1),
                                    labels = c("Not low income/access", "Low income/access"))

# Relevant numerical variables
num_vars <- c(
  "Pop2010", "OHU2010",
  "PovertyRate", "MedianFamilyIncome", 
  "TractLOWI",
  "TractKids", "TractSeniors",
  "TractWhite", "TractBlack", "TractAsian", "TractNHOPI", "TractAIAN","TractOMultir", "TractHispanic"
)
for (v in num_vars) all.data[[v]] <- as.numeric(all.data[[v]])


# compute populations as shares
lr.data <- all.data %>%
  mutate(
    TractLOWI_share = TractLOWI / Pop2010,
    TractKids_share = TractKids / Pop2010,
    TractSeniors_share = TractSeniors / Pop2010,
    TractWhite_share = TractWhite / Pop2010,
    TractBlack_share = TractBlack / Pop2010,
    TractAsian_share = TractAsian / Pop2010,
    TractNHOPI_share = TractNHOPI / Pop2010,
    TractAIAN_share = TractAIAN / Pop2010,
    TractOMultir_share = TractOMultir / Pop2010,
    TractHispanic_share = TractHispanic / Pop2010
  )

# compute shares as percentages
lr.data <- lr.data %>%
  mutate(
    TractLOWI_perc = TractLOWI_share * 100,
    TractKids_perc = TractKids_share * 100,
    TractSeniors_perc = TractSeniors_share * 100,
    TractWhite_perc = TractWhite_share * 100,
    TractBlack_perc = TractBlack_share * 100,
    TractAsian_perc = TractAsian_share * 100,
    TractNHOPI_perc = TractNHOPI_share * 100,
    TractAIAN_perc = TractAIAN_share * 100,
    TractOMultir_perc = TractOMultir_share * 100,
    TractHispanic_perc = TractHispanic_share * 100
  )

# compute logarithmic data
lr.data <- lr.data %>% mutate(
  l_TractSeniors_perc = log1p(TractSeniors_perc),
  l_TractKids_perc = log1p(TractKids_perc),
  l_TractWhite_perc = log1p(TractWhite_perc),
  l_TractBlack_perc = log1p(TractBlack_perc),
  l_TractHispanic_perc = log1p(TractHispanic_perc),
  l_TractAsian_perc = log1p(TractAsian_perc),
  l_TractNHOPI_perc = log1p(TractNHOPI_perc),
  l_TractAIAN_perc = log1p(TractAIAN_perc),
  l_TractOMultir_perc = log1p(TractOMultir_perc)
)

lr.data <- lr.data %>% 
  select(
    CensusTract, State, County,  # identifying info
    LILATracts_halfAnd10_f, LILATracts_halfAnd10, # response var
    # explanatory and supporting variables
    Pop2010,
    Urban, Urban_f,
    ## demographic as total count
    TractLOWI, 
    TractKids, TractSeniors, 
    TractWhite, TractBlack, TractAsian, TractNHOPI, TractAIAN, TractOMultir, TractHispanic, 
    # demographics as percentages (out of 100)
    TractKids_perc, TractSeniors_perc, 
    TractWhite_perc, TractBlack_perc, TractAsian_perc, TractNHOPI_perc, TractAIAN_perc, TractOMultir_perc, TractHispanic_perc,
    
    # percentage demographics on logarithmic scale
    l_TractKids_perc, l_TractSeniors_perc, 
    l_TractWhite_perc, l_TractBlack_perc, l_TractAsian_perc, l_TractNHOPI_perc, l_TractAIAN_perc, l_TractOMultir_perc, l_TractHispanic_perc
  ) %>% 
  filter(
    !is.na(TractKids), 
    !(State %in% c("Hawaii", "Alaska")), # Hawaii and Alaska are distinct (geographically and culturally) 
    Pop2010 >= 150 # removes most extreme skews (percentage-based)
  ) 

lr.data.clean <- filter(lr.data, !(CensusTract %in% c(
  "4001942700", "48365950500" # abnormally large  demographic %s, Hispanic / AIAN # TODO: how much influence do they have?
  ,"48341950200"
  )))

```

## Analytical techniques

A **logistic regression** model was fit. This model type is suitable for predicting information with two classes, such as whether a Census Tract is considered to have or not have a significant Low Income and Low Access population. 

Statistically significant coefficients of a logistic regression model model can describe in further detail the influence that different demographic information has on the status of the Census Tract. This is in contrast to other classification methods such as neural networks that and are difficult to interpret relationships from, and are slower to compute.


## Input data

The response variable is **Low Income and Low Access at 1/2 and 10 miles** (`LILATracts_halfAnd10`), measured as the odds of being flagged as low income and low access.

The following explanatory terms were included, measured on a logarithmic scale (see **data manipulation techniques**):

- The percentage of individuals in the Census Tract that are **children**, measured as less than 18 (`l_TractKids_perc`)
- The percentage of individuals in the Census Tract that are **Black** (`l_TractBlack_perc`)
- The percentage of individuals in the Census Tract that are **Asian** (`l_TractAsian_perc`)
- The percentage of individuals in the Census Tract that are **Hispanic** (`l_TractHispanic_perc`)
- The percentage of individuals in the Census Tract that are **Native Hawaiian or other Pacific Islander** (`l_TractNHOPI_perc`)
- The percentage of individuals in the Census Tract that are **American Indian (Native American) or Alaska Native** (`l_TractAIAN_perc`)
- The percentage of individuals in the Census Tract that are **other ethnicities not listed above** (excluding white), or identify as **multiple ethnicities** (`l_TractOMultir_perc`)

The interactions between whether the census tract was Urban or Rural (`Urban_f`), and a given ethnicity percentage, were included where found to be statistically significant (p-value < 0.001), with Urban as the reference level. This applied to **Hispanic**, **American Indian or Alaska Native**, and **Other ethnicities / Multiple Ethnicities**.

The only ethnicity measure not included was the percentage of the Census Tract that are **white**, which skewed in the opposite direction to all other ethnicities and was therefore treated as the reference level.

Similarly, there is high multi-collinearity between the percentage of individuals in a population that are children and the percentage that are seniors, as these are mutually exclusive categories (TODO: quantify).

Other models were fit that included interactions between the percentage of individuals that are children and the percentage of individuals that were various ethicities (Black, Hispanic, etc.). When only a selection of interactions with ethnicities were considered, the coefficients were found to be insignificant (for many, the p-value was > 0.05). A model was fit that incorporated all possible interaction terms; all of them were significant (p-value < 0.001) but the model was too difficult to interpret and coefficients were 2 or 3 orders of magnitude smaller.

TODO: appendix the model summary, reference it here

### Data manipulation techniques

For demographic explanatory terms (ethnicity and age-based continuous variables, such as **TractBlack**), total counts were converted to a logarithmically scaled percentage using the following equation: `log((TractBlack/Pop2010)*100 + 1)`.

Converting to percentages prevents the model fitting inadvertently to the population size. Many of the explanatory variables are skewed towards 0; taking the logarithm resulted in a better fitted model (AIC=65831, compared to AIC=69886 of a similar model fitted on the percentages without converting to the logarithmic scale).

### Constraining the scope and removing outliers

Census Tracts from the states of Hawaii and Alaska were removed. These data points were frequently found to be outliers, as their ethnicity profiles are skewed very differently to other states, due to geographical and historical reasons. Outcomes are thus only be considered applicable to states in the main land area that is North America.

Census Tracts where the total population was less than 150 were removed, affecting 253 Census Tracts. For Census Tracts with a population count at 150 or less, only 1 or 2 people are required to change the demographic by 1%. Values are more likely to be skewed, and results are difficult to interpret for these areas.

Lastly, three remaining rural Census Tracts were removed as outliers; one from Arizona and two from Texas. Their demographic patterns were distinct from other Census Tracts (for example, in the Census Tract from Arizona, 5444 out of 5628 individuals were Native American or Alaskan). 

TODO: quantify residuals for the outliers

```{r population_skew}
ggplot(all.data) +
  geom_violin(aes(x=Urban_f, y=Pop2010, fill=Urban_f)) +
  labs(y="Population count", x="Number of Census Tracts") + 
  scale_color_brewer(palette="Set1") +
  theme_minimal()
```
```{=latex}
\begin{center}
```
**Figure TODO**: Distribution of Population Count across all Census Tracts, split by Urban and Rural
```{=latex}
\end{center}
```

### Missing data

Tract Totals were missing for only 3 remaining Census Tracts; these records were removed. All 3 were Census Tracts in South Dakota, leaving 219 Census Tracts for this State. This is very not expected to bias the results.

Census Tracts missing the population count for low income and low access at 1/2 and 10 miles were **included** in this analysis; the response variable for all of these records is 0, as per the convention in the FARA dataset. Including these records prevents any potential bias (20.05% of records would be affected), and assumes that the decision to impute the corresponding flags as 0 was deliberate and appropriate. A further improvement on this research would be to attempt to gather information on the cause of these missing values. 

TODO: move improvement bit to conclusion

## Interpreting the model

```{r build_model}
model <- glm(
  LILATracts_halfAnd10_f ~
    l_TractKids_perc +
    l_TractAsian_perc 
      * Urban_f
  + l_TractHispanic_perc 
      * Urban_f
  + l_TractNHOPI_perc 
  + l_TractAIAN_perc 
      * Urban_f
  + l_TractOMultir_perc 
      * Urban_f
  + l_TractBlack_perc 

  , family="binomial"
  , data=lr.data.clean # outliers removed in data setup 
)

ci <- confint(model)
beta <- summary(model)$coeff
```

```{r make_hist_fun}
make_hist <- function(var, xlab, xmin=0, xmax=100) {
  ggplot(lr.data.clean, aes(x= {{ var }})) +
    geom_histogram() +
    labs(
      x=xlab,
      y= "Number of Census Tracts",
      title="Ethnicity percentages in Urban and Rural Census Tracts"
    ) +
    xlim(xmin, xmax) +
    facet_wrap(~Urban_f) +
    theme_minimal() +
    theme(
      panel.border=element_rect(colour="grey",size=0.2)
    )
}
```




# Conclusions and Recommendations

TODO


# References

Economic Research Service, U.S. Department of Agriculture. (2019).
*Food Access Research Atlas: Large Retailer Access Map (LRAM)
documentation*.
https://www.ers.usda.gov/data-products/food-access-research-atlas/documentation


# Appendix

```{r exp_beta}
exp(beta)
```

```{r exp_confint}
exp(ci)
```


