---
title: "Data471 Detailed Analysis Draft"
author: "Loren Whithair"
date: "`r Sys.Date()`"
output:
  pdf_document:
    md_extensions: -implicit_figures
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = "center")
```

```{=latex}
\setlength{\LTcapwidth}{\linewidth}
```

# Detailed Data Analysis

TODO: properly write up in report style
```{r detailed_analysis_libraries}
library(tidyr)
library(dplyr)
library(ggplot2)
library(car)
library(patchwork)
library(MASS)
```

## Analysis techniques used

> Explain what analysis techniques you used, and why they were suitable for the data and for the question you needed to answer.

- Logistic regression
- binary response, probability of being low income low access
- coefficients describe relationship between predictor var and response
- allows us to compare impact of different variables on binary yes/no is it low income low access


## Data manipulation

> Explain steps you took to manipulate the data, e.g. converting to logarithmic scale, standardizing continuous vars


- Urban converted to a factor, so that it will be treated as categorical
- Demographic info (TractBlack, TractKids etc,) did the following:
- converted to a percentage of total population (out of 100)
  - makes it easier to fit the model, each on similar scale (population varies wildly)
- took the logarithm of those percentages + 1 (log(x+1))
  - data is very skewed for most demographic info: e.g. % asian is typically around 10%. Makes sense conceptually to take log
  - this is why % white was not used
  - model fit on logged data had lower AIC than model fit on normal data (65k vs 69k), so seems like better fit
  - some values effectively 0, log 0 is undefined and log(x), x<1, is negative. Helps with the scaling

- certain data was removed:
  - remove Hawaii and Alaska: ethnicity skew completely different (geographical + cultural). Allow us to better predict for main area (TODO: COMMENT ON ETHICS)
  - remove those where Population < 150. Note if population < 100 then 1 person is 1%, hard to compare these small areas with much larger ones (10,000) TODO: comment on who this excludes, future research
  - a couple of remaining outliers with surprising demographic information (extremes based on skews). Some influence on data but not massive, found better fit without them. still plenty of data from those locations without them.

> Explain what you did to account for missing data

- those missing one key response var (TractKids), were missing ALL of them (TractBlack, TractWhite, etc) so just removed them (very small % of data, still have 70,000+ records)
- flags where the population equivalent is zero were included; allowed for more data on smaller areas. Data isn't technically "missing". Might make correlations + their p value weaker?


```{r detailed_analysis_setup}

#----------------------
set.seed(1899)

all.data <- read.csv(
  "data/2019/2019_Food_Access_Research_Atlas_Data/Food Access Research Atlas.csv",
  stringsAsFactors = FALSE,
  na.strings = c("NA", "NULL", "")
)

all.data <- as_tibble(all.data)

# Convert CensusTract to character:
all.data$CensusTract <- as.character(all.data$CensusTract)

# Convert Urban/Rural to factor, with Urban as reference
all.data$Urban_f <- factor(all.data$Urban, levels=c(1,0), labels = c("Urban", "Rural"))


# Response variable
all.data$LILATracts_halfAnd10_f <- factor(all.data$LILATracts_halfAnd10,
                                    levels=c(0,1),
                                    labels = c("Not low income/access", "Low income/access"))

# Relevant numerical variables
num_vars <- c(
  "Pop2010", "OHU2010",
  "PovertyRate", "MedianFamilyIncome", 
  "TractLOWI",
  "TractKids", "TractSeniors",
  "TractWhite", "TractBlack", "TractAsian", "TractNHOPI", "TractAIAN","TractOMultir", "TractHispanic"
)
for (v in num_vars) all.data[[v]] <- as.numeric(all.data[[v]])


# compute populations as shares
lr.data <- all.data %>%
  mutate(
    TractLOWI_share = TractLOWI / Pop2010,
    TractKids_share = TractKids / Pop2010,
    TractSeniors_share = TractSeniors / Pop2010,
    TractWhite_share = TractWhite / Pop2010,
    TractBlack_share = TractBlack / Pop2010,
    TractAsian_share = TractAsian / Pop2010,
    TractNHOPI_share = TractNHOPI / Pop2010,
    TractAIAN_share = TractAIAN / Pop2010,
    TractOMultir_share = TractOMultir / Pop2010,
    TractHispanic_share = TractHispanic / Pop2010
  )

# compute shares as percentages
lr.data <- lr.data %>%
  mutate(
    TractLOWI_perc = TractLOWI_share * 100,
    TractKids_perc = TractKids_share * 100,
    TractSeniors_perc = TractSeniors_share * 100,
    TractWhite_perc = TractWhite_share * 100,
    TractBlack_perc = TractBlack_share * 100,
    TractAsian_perc = TractAsian_share * 100,
    TractNHOPI_perc = TractNHOPI_share * 100,
    TractAIAN_perc = TractAIAN_share * 100,
    TractOMultir_perc = TractOMultir_share * 100,
    TractHispanic_perc = TractHispanic_share * 100
  )

# compute logarithmic data
lr.data <- lr.data %>% mutate(
  l_TractSeniors_perc = log1p(TractSeniors_perc),
  l_TractKids_perc = log1p(TractKids_perc),
  l_TractWhite_perc = log1p(TractWhite_perc),
  l_TractBlack_perc = log1p(TractBlack_perc),
  l_TractHispanic_perc = log1p(TractHispanic_perc),
  l_TractAsian_perc = log1p(TractAsian_perc),
  l_TractNHOPI_perc = log1p(TractNHOPI_perc),
  l_TractAIAN_perc = log1p(TractAIAN_perc),
  l_TractOMultir_perc = log1p(TractOMultir_perc)
)

lr.data <- lr.data %>% 
  select(
    CensusTract, State, County,  # identifying info
    LILATracts_halfAnd10_f, LILATracts_halfAnd10, # response var
    # explanatory and supporting variables
    Pop2010,
    Urban, Urban_f,
    ## demographic as total count
    TractLOWI, 
    TractKids, TractSeniors, 
    TractWhite, TractBlack, TractAsian, TractNHOPI, TractAIAN, TractOMultir, TractHispanic, 
    # demographics as percentages (out of 100)
    TractKids_perc, TractSeniors_perc, 
    TractWhite_perc, TractBlack_perc, TractAsian_perc, TractNHOPI_perc, TractAIAN_perc, TractOMultir_perc, TractHispanic_perc,
    
    # percentage demographics on logarithmic scale
    l_TractKids_perc, l_TractSeniors_perc, 
    l_TractWhite_perc, l_TractBlack_perc, l_TractAsian_perc, l_TractNHOPI_perc, l_TractAIAN_perc, l_TractOMultir_perc, l_TractHispanic_perc
  ) %>% 
  filter(
    !is.na(TractKids), 
    !(State %in% c("Hawaii", "Alaska")), # Hawaii and Alaska are distinct (geographically and culturally) 
    Pop2010 >= 150 # removes most extreme skews (percentage-based)
  ) 

lr.data.clean <- filter(lr.data, !(CensusTract %in% c(
  "4001942700", "48365950500" # abnormally large demographic %s, Hispanic / AIAN # TODO: how much influence do they have?
  # ,"48341950200"
  )))

lr.data.clean <- lr.data
```


## The model
> Specify the full list of terms included in any model you fit, i.e. list the features or explanatory variables and any transformations you applied to them or any interactions between them


The response variable is **Low Income, Low Access, at 1/2 and 10 miles (LILATractshalfAnd10)*. It is a flag for if the Census Tract is low access and a significant number of people in the census tract (at least 500 people, or at least 33%) are more than 1/2 mile or 10 miles from a supermarket, for Urban and Rural respectively.

Response variables that are taken as % of total pop, converted to log scale:

- %kids
- %asian
- %hispanic
- %nhopi (native hawaiian and other pasific islander)
- %aian (american indian and alaska native)
- %omultir (other races, or identifying as multiple races)
- %black

The flag for whether the Cesus Tract is Urban or Rural was included as a response too.

More detailed information on measurement methods can be found in the Large Retail Access Map Reference Guide (2).

Choices made:
- used %white as reference level
  - largest % in most CensusTracts, so skewed the other way to all other vars. Model was more interpretable without it
- %kids focussed on rather than %seniors; not exhaustive but inversely proportional. Variance inflation factor was very high for %seniors, so picked %kids
- including interaction terms between age + ethnicity (% kids and TractBlack) were tried; either coefficients were insignificant, or only significant if done on ALL ethnicity combinations + Urban/Rural. Model too large to interpret and effects too small to gain insight
- just looking at TractKids didn't provide much insight; focussed on ethnicity demographics
- Start off with all ethnicities, Urban/Rural, and interaction terms between all ethnicities and urban/rural. The Urban/Rural interaction with ethnicities that were insignificant (p > 0.001)



```{r model_build}
# commented out interaction terms were insignificant
model <- glm(
  LILATracts_halfAnd10_f ~
    l_TractKids_perc +
    l_TractAsian_perc 
      * Urban_f
  + l_TractHispanic_perc 
      * Urban_f
  + l_TractNHOPI_perc 
  + l_TractAIAN_perc 
      * Urban_f
  + l_TractOMultir_perc 
      * Urban_f
  + l_TractBlack_perc 

  , family="binomial"
  , data=lr.data.clean # outliers removed in data setup 
)

summary(model)
```

```{r outliers}
influencePlot(model)
```
```{r outliers_x}
lr.data.clean[c(64301, 64693),]
# lr.data.clean[c(64300),] # 48341950200, Texas, Moore County

studres(model)

```

```{r model_fit}
summary(model)
```


```{r resid}
plot(model, which=1)
```
## Residuals and model fit

- model is the right type of fit: see that the correlations between input and response vars are linear (figure below)

```{r fitted_vs_inputs}
p1 <- ggplot(lr.out, 
       aes(
         x=l_TractBlack_perc,
         y=model.fitted
       )) +
  geom_point(
    alpha=0.1,
    colour=
  ) + labs(
    x="Percentage of tract that is black (natural log of adjusted value)",
    y="Fitted model value"
  ) +
  geom_abline(slope= 0.15,
              intercept=0.0,
              colour="black",
              size=1
  ) + theme_minimal() +
  scale_colour_brewer(palette = "Set1")

p2 <- ggplot(
  lr.out, aes(x=l_TractHispanic_perc, y=model.fitted, colour=Urban_f
  )) +
  geom_point(
    alpha=0.1
  ) + labs(
    x="Percentage of tract that is Hispanic (natural log of adjusted value)",
    y="Fitted model value",
    colour="Population Density type"
  ) + 
  geom_abline(slope= 0.05,
              intercept=0.1,
              colour="black",
              size=1
  ) + 
  geom_abline(slope= 0.001,
              intercept=0.05,
              colour="black",
              size=1
  ) +
  theme_minimal() + 
  scale_colour_brewer(palette = "Set1") +
  guides(colour = guide_legend(override.aes = list(alpha = 1)))

p1 + p2
```

```{=latex}
\begin{center}
```
**Figure TODO**: Comparisons of inputs and fitted values of the model.  
```{=latex}
\end{center}
```


## Input var coefficients

```{r hist_fn}
make_hist <- function(var, xlab, xmin=0, xmax=100) {
  ggplot(lr.data.clean, aes(x= {{ var }})) +
    geom_histogram() +
    labs(
      x=xlab,
      y= "Number of Census Tracts",
      title="Ethnicity percentages in Urban and Rural Census Tracts"
    ) +
    xlim(xmin, xmax) +
    facet_wrap(~Urban_f) +
    theme_minimal() +
    theme(
      panel.border=element_rect(colour="grey",size=0.2)
    )
    
}

# # test
make_hist(TractBlack_perc, "% Black")
```

```{r coef_setup}
ci <- confint(model) # must take exponent of this too!
beta <- summary(model)$coeff
```


```{r coef_analysis}
summary(model)
```

```{r coef_analysis}
exp(beta)
```


**Note on log increases**:
used natural log of percentage. Means that increases at lower values have larger impact than increase at higher values
(e.g. log(2) - log(1) == log(20) - log(10))
An increase from 10% to 20% is roughly equal in impact to an increase from 1% to 2%

Also, number of actual people depends on total population size.

```{r pop_violin}
ggplot(lr.data.clean) +
  geom_violin(aes(x=Urban_f, y=Pop2010)) + 
  labs(
    title= "Frequency of different Population Sizes among Census Tracts",
    x="Population of Census Tract",
    y="Frequency"
  ) +
  theme_minimal() 
```



### General:

- urban_f not interesting on it's own
- tractblack worse in both areas (no interaction w urban+rural)
- tractAIAN bad in urban, worse 

**Urban_f**

Urban vs rural: rural is less likely to be LILA (62% to 48% less likely, 95% CI, p-value = 2e-16).
This could  be a consequence of measurement (different distances). More interesting is Urban/Rural as interaction term.


**l_TractKids_perc**

Odds decrease with more children in the population, 23% less per log unit increase.
However cannot say anything in more detail; didn't make parsimonious model with low enough p-value for interaction terms w %kids.

We don't know what % of those children are in each demographic (e.g. 70% White, 10% children, doesn't necessarily mean 7% white children)
Potential for further analysis

**Tract Asian perc**: 

main effect is negative
- 63% less likely for 1 unit log increase in urban areas (1-2, or 10-20)
- In rural areas, even greater impact: 0.37 * 0.24 = 0.087, 91.% less likely per 1 unit log increase

**TractBlack_perc**

1.5x more likely to be low income low acces for each log unit increase of TractBlack (95% CI (1.47, 1.52), p-value 2e-16).

No significant difference found between Urban and Rural areas for this effect (p-value too low). Cannot say there's a difference.

In other words, odds increase in similar fashion in both types of region. 

TODO: comment on scale, how many ppl this would actually be. 
TODO: typical range?

**Tract Hispanic perc**

1.6x LESS likely in urban?

40% less likely in urban areas per hispanic % unit log increase

1.3x more likely in rural areas per hispanic % unit log increase


Urban: 0.6x as likely (goes down)
Rural: 1.3x as likely (goes up) odds are DOUBLED in rural area compared to 

OR can explain as:
40% less likely
30% more likely


Why different direction?

Something to do with skew urban vs rural:

```{r hispanic_boxplots}
ggplot(lr.data.clean) +
  geom_boxplot(aes(y=TractHispanic_perc, x=Urban_f)) +
  # facet_wrap(~Urban_f) +
  theme_minimal()
```

Potentially due to skew of data urban vs rural? Could be reflection of 

**TractOMultir_perc**

Largest impact: TractOMultir (multi-racial) had biggest effect: 3.3x (TODO: CI, p-value) more likely per 1 unit log increase.

inverse effect in rural areas: 0.8x as likely (i.e. 20% less likely). Howeve

encompasses "other" races (i.e. not White, black, native american, asian, hawaiian, pacific islander) AND multi-racial.

Cannot say which has the biggest effect, cannot say if distribution is different between urban and rural. limit interpretation

Can say they are a minority (non-white, not exclusively white) and that this significantly increases likelihood in Urban areas, slightly reduces in rural.

Change in direction could be due to variety of things: the actual races / split between other and multi may be different in these areas.
See boxplot below: skewed way further to 0 for rural than urban, could be consequence of logarithmic scaling (don't see effects )

i.e. wouldn't say it's a representation of a real-world phenomenon, more probably a consequence of the model.

TODO: comment on actual values, # of people needed


```{r omultir_boxplot}
ggplot(lr.data.clean) +
  geom_boxplot(aes(y=TractOMultir_perc, x=Urban_f)) +
  scale_color_brewer(palette="Set1") +
  theme_minimal()
```
```{r omultir_boxplot_totals}
ggplot(lr.data.clean) +
  geom_boxplot(aes(y=TractOMultir, x=Urban_f)) +
  scale_color_brewer(palette="Set1") +
  theme_minimal()
```


```{r barplot_urban_rural}
ggplot(lr.data.clean) +
  geom_bar(aes(x=Urban_f), fill="tomato") +
  labs(x="Population Density Type", y="Number of Census Tracts", title="Urban vs Rural Census Tracts") +
  theme_minimal()
```


```{r omultir_stats}
lr.data.clean %>%
  group_by(Urban_f) %>%
  summarise(
    mean_pop = mean(Pop2010),
    mean_multi_perc= mean(TractOMultir_perc),
    min_multi_perc= min(TractOMultir_perc),
    max_multi_perc= max(TractOMultir_perc),
    
    mean_multi= mean(TractOMultir),
    min_multi= min(TractOMultir),
    max_multi= max(TractOMultir)
  )
```


```{r omultir_hist}
make_hist(TractOMultir_perc, "% Other / Multiple Races", xmax=60)
```

**TractAIAN_perc** (native american, alaskan)

1.51x more likely per % logged in urban areas (CI!) 

Even worse in rural: extra 1.3224789 x more likely (total of 1.9x more likely)

```{r laomultir}

p1 <- all.data %>%
  filter(Urban_f =="Urban") %>%
  mutate(
    # la_omulti_perc = safe_divide(laomultir1, TractOMultir)
    la_omulti_perc = laomultir1/TractOMultir
    ) %>%
  ggplot() +
  geom_histogram(aes(x=la_omulti_perc)) + 
  # xlim(0,100) +
    labs(title="Urban", y="% of omultir that are LA")


p2 <- all.data %>%
  filter(Urban_f =="Rural") %>%
  mutate(
    # la_omulti_perc = safe_divide(laomultir1, TractOMultir)
    la_omulti_perc = laomultir1/TractOMultir
    ) %>%
  ggplot() +
  geom_histogram(aes(x=la_omulti_perc)) + 
  # xlim(0,100) +
    labs(title="Rural", y="% of omultir that are LA")

p1 + p2
```




```{r compare_ranges}
print("Range of % black:")
print(range(lr.data.clean$TractBlack_perc))
print("Range of % multiracial: ")
print(range(lr.data.clean$TractOMultir_perc))
```

```{r hist_black}
hist(lr.data.clean$TractBlack)
```

```{r hist_multi}
ggplot(lr.data.clean)+
  geom_histogram(aes(x=TractOMultir)) +facet_wrap(~Urban_f) + theme_minimal()
# hist(lr.data.clean$TractOMultir)
```

### confidence intervals

```{r ci_analysis}
exp(ci)
```

# Conclusions and Recommendations

- Conclusion: black, 


# References

(1) Economic Research Service, U.S. Department of Agriculture. (2019).
*Food Access Research Atlas: Large Retailer Access Map (LRAM) documentation*.
https://www.ers.usda.gov/data-products/food-access-research-atlas/documentation


(2) *Food Access Research Atlas - Large Retailer Access Map Reference Guide*
https://www.ers.usda.gov/data-products/food-access-research-atlas/documentation/large-retailer-access-map-reference-guide
